signalHsmm - a novel semi-Markov model of eukaryotic signal peptides

### 2015-06-16 paper.tex
first draft of HMM section

@@

\begin{figure}[ht]\centering
\includegraphics[width=0.9\textwidth]{reglen.png}
\caption{Distribution of the lengths of signal peptide regions. The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}.
\label{fig:reglen}
\end{figure}

@@

Tu troche matematyki o ukrytych modelach Markowa~\citep{1989rabinera}.

>>>>

% rysunki
\usepackage{tikz}
\usepackage{ifthen}
\usepackage{xxcolor}
\usetikzlibrary{arrows}
\usetikzlibrary[topaths]
\usetikzlibrary{decorations.pathreplacing}

@@

\begin{figure}[ht]\centering
\includegraphics[width=0.9\textwidth]{reglen.png}
\caption{A) Distribution of lengths of signal peptides. B) Distribution of the lengths of signal peptide regions. The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}.
\label{fig:reglen}
\end{figure}

@@

Hidden Markov Model(HMM) is a widely used machine learning technique which, in contrast to black-box methods like neural network used by SignalP, 
is easy to interpret and allows researcher to analyze and understand structure of modeled phenomenon. 
In case of signal-peptide, we might wish to verify whether regions really describe it. Also we might capture
structure for some special subgroups and with HMM we can verify biological hypothesis.

Suppose you have a sequence of observations e.g. amino acids. What is of interest is
understanding underlying cause of their occurrence. 
HMM aims to answer that question by assuming a specific yet flexible structure of the problem.
HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
These are "the cause" for the observations. At every step, hidden state might change according to transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
Second process $E_{t=1}^T$ is observation process, defined on the set of possible observations $\{O_1, \dots, O_m\}$. 
Observations are assumed to occur independently. Their distribution depends only on hidden state which emits it. 
Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, observations are (degenerated) amino acids.
Main goal in signalHsmm is to find, for a given peptide, most probable regions boundaries. This is achieved with
Viterbi algorithm.
For a good reference on HMM see~\citep{1989rabinera}.

In regular HMM, hidden state duration -- number of observations emitted by the hidden state, has geometric distribution.
\citep{Durbin98biologicalsequence} shows how to extend it for some different distributions without significant increase in computational 
complexity. Similar ideas were used for signal peptide recognition, for example in~\citep{2004klla}. 
However it still not flexible enough. Empirical regions length distributions (see Figure~\ref{fig:reglen}),
is difficult to capture in that way.

%rysunek pokazujący o co chodzi w HMM
\begin{figure}[h]
\centering
\tikzstyle{block} = [draw,shape=circle, top color=white, bottom color=green!50!red!70 ,minimum size=4em]
\tikzstyle{blockComp} = [draw,shape=circle, top color=white, bottom color=red!50!black!70 ,minimum size=2.5em]
% diameter of semicircle used to indicate that two lines are not connected
\def\radius{.7mm} 
\tikzstyle{branch}=[shape=rectangle, top color=white, bottom color=green!50!red!70 ,minimum size=3pt,inner sep=0pt]
\tikzstyle{branch2}=[shape=rectangle, top color=white, bottom color=red!50!black!70 ,minimum size=3pt,inner sep=0pt]
\def\n{11}
 \tikzstyle{line} = [draw=black, color=black!30!white!50, line width=1.5mm, -latex'] 
 \tikzstyle{line2} = [draw=black, color=red!30!black!50, line width=1.5mm, -latex']
\tikzstyle{frame}=[text=black,above, bottom color=white, top color=blue!50!black!70  ]
\tikzstyle{frameComp}=[text=black,below, bottom color=white, top color=red!50!black!70  ]
 \tikzstyle{line3} = [color=green!50!red!70, line width=1.5mm]
\def\names{{"$O_{1,1}$","...","$O_{1,d_1}$", "...","$O_{k, 1}$", "...", "$O_{k, d_k}$"}}%
\begin{tikzpicture}[>=latex']
%wąsate nawiasy
\draw [decorate, decoration={brace,amplitude=10pt},line3,xshift=0pt,yshift=0pt] (2.5,1) -> (10.5,1) node [branch,midway,yshift=0.8cm,color=black] 
	{\textbf{Hidden states}};
% \draw [->,decoration={brace,mirror,amplitude=10pt},line2,xshift=0pt,yshift=0pt] (11,-3) -- (3,-3) node [branch2,midway,yshift=-0.6cm,color=black] 
% 	{\textbf{Kierunek odczytu na nici Cricka ($3' \to 5'$)}}; 
   % Draw blocks, inputs and outputs

\node[block, text width=3em] at (1+2,0) (block1) {\footnotesize 1st hidden state};
\node[blockComp] at (-.5+2,-2)  (komp1) {\footnotesize \pgfmathparse{\names[0]}\pgfmathresult};
\node[blockComp] at (1+2,-2)  (komp2) {\footnotesize \pgfmathparse{\names[1]}\pgfmathresult};
\node[blockComp] at (2.5+2,-2)  (komp3) {\footnotesize \pgfmathparse{\names[2]}\pgfmathresult};
\draw[line,->]  (block1) --  (komp1);
\draw[line,->]  (block1) --  (komp2);
\draw[line,->]  (block1) --  (komp3);
\draw[line3,->] (block1.east) -- +(2,0) node [branch,midway,yshift=0.2cm,xshift=-0.2cm,color=black] {\footnotesize \textbf{transition}};

\node[block, bottom color=green!50!red!80,] at (4.5+2,0) (block2) {...};
\draw[line3,->] (block2.east) -- +(2,0) node [branch,midway,yshift=0.2cm,xshift=-0.2cm,color=black] {\footnotesize \textbf{transition}};
\node[blockComp] at (4.5+2,-2)  (komp4) {\footnotesize \pgfmathparse{\names[3]}\pgfmathresult};
\draw[line,->]  (block2) --  (komp4) ;

\node[block,text width=3.5em, bottom color=green!50!red!90] at (8+2,0) (block3) {\footnotesize $k^{th}$ hidden state};
\node[blockComp] at (6.5+2,-2)  (komp5) {\footnotesize \pgfmathparse{\names[4]}\pgfmathresult};
\node[blockComp] at (8+2,-2)  (komp6) {\footnotesize \pgfmathparse{\names[5]}\pgfmathresult};
\node[blockComp] at (9.5+2,-2)  (komp7) {\footnotesize \pgfmathparse{\names[6]}\pgfmathresult};
\draw[line,->]  (block3) --  (komp5);
\draw[line,->]  (block3) --  (komp6);
\draw[line,->]  (block3) --  (komp7);

% \draw[<->, color=red!30!black!50, line width=1.5mm] (komp1) |- +(5,-1.5) node[line2, yshift=-0.5cm] {\textbf{Observations}} -| (komp7);

\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=20pt,mirror,raise=10pt},yshift=-10pt]
(1,-1.8) -- (12,-1.8) node [branch2, midway, yshift=-1.5cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};

\end{tikzpicture}
\caption{Hidden semi-Markov Model scheme}
\label{fig:hsmm}
\end{figure}

The model we use is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing 
any given hidden state length distribution. Scheme is shown in figure~\ref{fig:hsmm}.
Model is given by matrices $A$ and $B$ as before, as well as probabilities of duration length in hidden states.
$$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
$D$ is maximal allowed duration.
As our datasets are of reasonable size, $D$ is small -- around 30, computational effort is not much 
higher than in the regular HMM.

The gain is not only a better fit to the data but also, as number of hidden states is smaller, easier interpretation 
for researcher without mathematical background.

HSMM can be further extend to use some specific knowledge of motives - n-grams.
